---
title: "simulate-survival"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simulate-survival}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(survivaltools)
library(data.table)
library(parallel)
library(survival)
set.seed(20)
```


## Data generating functions

### generating participant data

generate_ppt_table is a simple function to randomly generate age at enrollment, date of study entry, and the last possible day of follow-up for a given number of participants:
```{r}
 generate_ppt_table(1000)
```
Note that the units are always in days since "1970-01-01". It's just easier to work with integers than dates. 

Every time this function is called, it generates a new set of participant data: 
```{r}
 generate_ppt_table(1000)
```


This is a typical format for representing survival data (one row per participant), but for simulations we need to expand this dataset such that there's a row for every time point in every participant:

```{r}
ppt_tbl <- generate_ppt_table(1000)
ppt_tbl_expanded <- expand_ppt_table(ppt_tbl)
```
The result is fully crossed between all possible time points and every participant, but not every participant is actually at risk at all of those time points. the at_risk column indicates whether `time` is in a participant's interval `(t_start,t_exit]` (notice the open interval on the left. this is the convention with cox models that a participant is not at risk on day of entry)


### the baseline hazard
A quick reminder of how baseline hazard works for a cox model:

```{r}
baseline_hazard=function(x,ppt){0.000003}
hazard <- function(x){baseline_hazard(x)*exp((1/20000)*x)}
hazard(365*70)/hazard(365*50)
 #20 years of aging is a HR of ~1.44=exp(20*365*1/20000)
```

baseline_hazard $h_0(x)$ and hazard $h(x|X)$ represent functions which vary over time ($x$). The hazard function is just the baseline hazard times a transformed linear function of the covariates $X$: 
$h(x|X)=h_0(x)*exp(X\beta)$

We could use a constant hazard in our simulations, but to do more realistic simulations that deal with time-varying data, it's probably better to generate a hazard function that isn't constant over time and is different for each simulation run.

I've written an R function which generates a baseline hazards function (using a spline fit to randomly generated points):


```{r} 
times <- ppt_tbl_expanded[ppt_id==1,time] #extract unique times
baseline_hazards <- generate_baseline_hazard(times)
plot(times,baseline_hazards,type="l",ylim=c(2e-6,3.5e-6))
```

The hazard function decreases over time by design. This makes it easy to induce
correlations with secular trends in confounding time-varying.

`generate_baseline_hazard` will generate a new set of baseline hazards every time it's called. Here we plot several different baseline hazards:

```{r}
hazard_matrix <- replicate(6,generate_baseline_hazard(times))
matplot(times,hazard_matrix,col=1:ncol(hazard_matrix),type="l",lty=1)
```

Note that these hazard occur over a very tight range around ~0.000003 (the default for generate_hazard_table) with a strong linear slope. This corresponds to a reasonable hazard (or conditional probability) when the x axis is units of days.

### time-varying variables:  exposure

In addition to generating participant recruitment timing information and the baseline hazard function, we also need a function to generate time-varying covariates such as PM2.5.

glossing over the details, the result is a set of participant-specific exposure curves that have a linear trend over time and are built on the same basis functions with random intercepts.

In the following plot, each color is a ppt and each panel is a different set of randomly generated exposure functions.

```{r,fig.height = 8, fig.width=8, fig.align='center'}
par(mfrow=c(2,2))

for(j in 1:4){
  exposures <- generate_exposure_matrix(times,1:10,degree=10)
  matplot(times,exposures,col=1:ncol(exposures),type="l",lty=1)
}
par(mfrow=c(1,1))
```



# Survival simulations



A simple simulation with calendar time as the baseline time-scale.
No pm2.5 exposure, but include an effect of aging and an effect of smoking.
We shouldn't expect to need to adjust for age here because even though age determines
risk it's not associated with smoking status.
```{r}

results <- lapply(X=1:2,FUN=function(x){
  library(survivaltools)
  library(survival)
  
  ppt_tbl1 <- generate_ppt_table(1000)
  ppt_tbl1[, smoking:=rbinom(.N,size=1,prob=.5)]
  x <- prep_tv_dataset(ppt_tbl1,include_exposure = FALSE)
  
  
  q <- simulate_survival(x=x,
                         covar_names=c("smoking","age_since65_tv"),
                         covar_coefs=c(1.2,1/10000)
  )
  
  q <- q[ppt_tbl1,on=c("ppt_id","t_start")]
  
  
  q[,baseline_age_c:=floor(baseline_age/365.25)]
  
  
  out <- list(
    extract_coef_table(coxph(Surv(t_start,t_end,event)~strata(baseline_age_c)+smoking,data=q)),
    extract_coef_table(coxph(Surv(t_start,t_end,event)~baseline_age_c+smoking,data=q)),
    extract_coef_table(coxph(Surv(t_start,t_end,event)~smoking,data=q))
  )
  
  attr(out, "n_events") <- q[, sum(event)]
  out
  
}) 

```


```{r,eval=FALSE}
summary(sapply(results, attr, "n_events"))
```

```{r,eval=FALSE}
summarize_simulation_table(rbindlist(unlist(results,recursive=FALSE)))
```

As expected, results are unbiased regardless of whether we stratify, adjust, or ignore age.




## Survival simulation with a time-varying variable. model on calendar time-scale

Simulation:

Add two time-varying covariates:  time-varying "exposure" which decreases over time, time-varying age which obviously increases over time. Include non-zero coefficients for these terms. Simulate baseline hazard on a baseline timescale of calendar time with a temporal trend in that baseline time-scale (hazard decreases over time). 


Model:
Baseline time-scale is calendar time (exact). Consider various adjustments/stratification for age discretized to year.

- no adjustment for age
- adjust for baseline age
- adjust for time-varying age
- adjust for time-varying age interacted with baseline year
- stratify by baseline age
- stratify by time-varying age



```{r}



results2 <- lapply(X=1:2,FUN=function(x){
  library(survivaltools)
  library(survival)
  
  
  ppt_tbl1 <- generate_ppt_table(1000)
  x <- prep_tv_dataset(ppt_tbl1)
  

  q <- simulate_survival(x=x,
                         covar_names=c("age_since65_tv","exposure"),
                         covar_coefs=c(1/10000,0.01)
 )

  q <- q[ppt_tbl1,on=c("ppt_id","t_start")]
  
  
  q[,baseline_age_c:=floor(baseline_age/365.25)]
  q[,baseline_year:=1970L+floor(t_start/365.25)]
  
  q_split <- split_by_event_times(q,interval_vars=c("t_start","t_end"),id_var="ppt_id",event_indicator="event")
  
  q_split[x, exposure:=exposure, on=c("ppt_id",end="time")]
  
  q_split[, tv_age_c:=baseline_age_c+floor((end-t_start)/365.25)]
  
  

  
  formulas <- list(
       Surv(start,end,event)~strata(baseline_age_c)+exposure,
       Surv(start,end,event)~baseline_age_c+exposure,
       Surv(start,end,event)~baseline_age_c*baseline_year+exposure,
       Surv(start,end,event)~exposure,
       Surv(start,end,event)~strata(tv_age_c)+exposure,
       Surv(start,end,event)~tv_age_c+exposure,
       Surv(start,end,event)~tv_age_c*baseline_year+exposure,
       Surv(start,end,event)~exposure
       )
  
  out <- future.apply::future_lapply(formulas, function(x){
    extract_coef_table(coxph(x,data=q_split),RHS=as.character(x)[3])
  })
  
  attr(out, "n_events") <- q[, sum(event)]
  out
  
}) 


```


```{r,eval=FALSE}
summary(sapply(results2, attr, "n_events"))
```

```{r,eval=FALSE}
summarize_simulation_table(rbindlist(unlist(results2,recursive=FALSE)))
```
